---
title: "Trach Study Cohort Selection"
author: "HST953"
date: "4/18/2019"
output: html_document
---

### Utility Functions

#### rename()

```{r}
rename <- function(dat, avoid_names, prefix){
        for (name in colnames(dat)[!(colnames(dat) %in% avoid_names)]){
            colnames(dat)[which(colnames(dat) == name)] <- paste(prefix, name, sep = '')
        }
    return(dat)
}
```

#### ventCheck()

```{r}
ventCheck <- function(dat, days, admission_level, buffer){
    #Generate some temporary variables
    tmp <- data.frame() 
    res <- data.frame()
    
    ## Order
    dat <- dat[with(dat, order(SUBJECT_ID, HADM_ID, ICUSTAY_ID, ADMITTIME, INTIME, STARTTIME, ENDTIME)),]
      
    #For each unique icustay in the data set 
    for (id in unique(dat[[admission_level]])){
        
      ## Subset subject data 
      tmp <- dat[(dat[[admission_level]] == id),] 
      
      ## Elevator Music
      ## message("ID: ", id)
      
      ## Time check here-- check to see that the latest endtime is greater than the earliest starttime PLUS some time period
      if (((as.POSIXlt(as.character(tmp[nrow(tmp),]$ENDTIME), tz = "EST")) + (buffer * (24 * 60 * 60))) >= (as.POSIXlt(as.character(tmp[1,]$STARTTIME), tz = "EST") + (days * (24 * 60 * 60) ))){
          
          ## First start time in the block
          tmp$FIRST_VENT_STARTTIME <- rep(tmp$STARTTIME[1], each = nrow(tmp))
          
          ## More elevator music
          message("Included ID: ", tmp[[admission_level]][1])
          ## Bind results
          res <- rbind(res, tmp)
      }

    }
    rownames(res) <- 1:nrow(res)

    return(res)
}
```

#### print_info()

```{r}
print_info <- function(dat){
    print(length(unique(dat$SUBJECT_ID)))
    print(length(unique(dat$HADM_ID)))
    print(length(unique(dat$ICUSTAY_ID)))
}
```

# Demographic and Cohort Data

## Load Admission Data & Initial Cleaning

Load `ADMISSIONS`, `PATIENTS`, and `ICUSTAYS` for cohort data. Load `icustay_detail` for more derived data. __Note:__ `icustay_detail` data are derived from `ICUSTAYS`

```{r}
## Load ADMISSIONS Table to join on HADM_ID
adm <- read.csv("~/MIMIC-III/ADMISSIONS.csv", 
                header = T, stringsAsFactors = F)

## Load PATIENTS Table to join on SUBJECT_ID
pat <- read.csv("~/MIMIC-III/PATIENTS.csv", 
                header = T, stringsAsFactors = F)

## Load ICUSTAYS Table to join on SUBJECT_ID, HADM_ID
stays <- read.csv("~/MIMIC-III/ICUSTAYS.csv", 
                  header = T, stringsAsFactors = F)

## Load icustay_detail to join with ICUSTAYS
## Note: icustay_detail is a derived table
stays_detail <- read.csv("~/MIMIC-III/icustay_detail.csv", 
                         header = T, stringsAsFactors = F)

## Convert colnames to uppercase
colnames(stays_detail) <- toupper(colnames(stays_detail))
```

## Initial Cohort (Derived from ICUSTAYS)

As all patients must be acutely ill, we will begin at the ICU level.

```{r}
cat("There are", length(unique(stays$SUBJECT_ID)), "unique patients in this cohort.\n")
cat("There are", length(unique(stays$HADM_ID)), "unique hospital admissions associated with this cohort.\n")
cat("There are", length(unique(stays$ICUSTAY_ID)), "unique ICUSTAYs in this cohort.\n")

cat(length(unique(stays$ICUSTAY_ID)) - length(unique(stays_detail$ICUSTAY_ID)), "ICUSTAYs are missing CHARTEVENTS data.\n")
cat(length(unique(stays$HADM_ID)) - length(unique(stays_detail$HADM_ID)), "HOSPITAL ADMISSIONS are missing CHARTEVENTS data.\n")
cat(length(unique(stays$SUBJECT_ID)) - length(unique(stays_detail$SUBJECT_ID)), "Patients were removed for having no CHARTEVENTS data.\n")
```

### Merging Admission, Patient, and ICU Data

1. Remove `ROW_ID`, which is an index peculiar to each table.
2. Join `ICUSTAYS` to derived `ICUSTAYS_DETAIL` table on tabular intersection. `ICUSTAYS` contains information that the derived `ICUSTAYS_DETAIL` does not.
3. Join `ADMISSIONS` to `PATIENTS` on `SUBJECT_ID` for `GENDER`, `DOB`, `DOD` info.
4. Join `ICUSTAYS` to `PATIENTS` and `ADMISSIONS` on tabular intersection.

```{r}
## (1)
adm$ROW_ID <- NULL
pat$ROW_ID <- NULL
stays$ROW_ID <- NULL

## (2)
stays <- merge(stays, stays_detail, by = intersect(colnames(stays), colnames(stays_detail)))

## Clean stays_detail
rm(stays_detail)

## (3)
dat <- merge(adm, pat, by = "SUBJECT_ID")
dim(dat)

## Clean
rm(adm, pat)

## (4)
dat <- merge(dat, stays, by = intersect(colnames(dat), colnames(stays)))
dim(dat)

## Clean
rm(stays)

## Look again for pt, hadm, icustay numbers
cat("There are", length(unique(dat$SUBJECT_ID)), "unique patients in this cohort.\n")
cat("There are", length(unique(dat$HADM_ID)), "unique hospital admissions in this cohort.\n")
cat("There are", length(unique(dat$ICUSTAY_ID)), "unique ICUSTAYs in this cohort.\n")
```

### Correct Ages and Remove patients younger than 18 years of age

```{r}
## Correct nonegenarian ages by imputing median age (91.4)
dat[(dat$ADMISSION_AGE >= 90), ]$ADMISSION_AGE <- 91.4

## Remove patients youner than 18
tmp <- dat[(dat$ADMISSION_AGE >= 18), ]

cat(length(unique(dat$SUBJECT_ID)) - length(unique(tmp$SUBJECT_ID)), "patients below the age of 18 removed.\n")
cat(length(unique(dat$HADM_ID)) - length(unique(tmp$HADM_ID)), "hospital admissions for patients below the age of 18 removed.\n")
cat(length(unique(dat$ICUSTAY_ID)) - length(unique(tmp$ICUSTAY_ID)), "ICU stays for patients below the age of 18 removed.\n")

## dat gets tmp, remove tmp
dat <- tmp
rm(tmp)

cat("There are", length(unique(dat$SUBJECT_ID)), "unique patients in this cohort.\n")
cat("There are", length(unique(dat$HADM_ID)), "unique hospital admissions in this cohort.\n")
cat("There are", length(unique(dat$ICUSTAY_ID)), "unique ICUSTAYs in this cohort.\n")
```

## Diagnoses & Procedures

Load `DIAGNOSES_ICD` and `D_ICD_DIAGNOSES`.

```{r}
## Load ICD codes
icd <- merge(read.csv("~/MIMIC-III/DIAGNOSES_ICD.csv", header = T, stringsAsFactors = F), 
             read.csv("~/MIMIC-III/D_ICD_DIAGNOSES.csv", header = T, stringsAsFactors = F), by = "ICD9_CODE")

## Clean Row IDs
icd$ROW_ID.x <- NULL
icd$ROW_ID.y <- NULL

## Rename columns to tag them with diagnosis for regerence
icd <- rename(icd, c("SUBJECT_ID", "HADM_ID"), "DIAGNOSIS_")

## View head
head(icd)
```

### Malignant Neoplasm of Head/Face/Neck

Look for `14XX` ICDs indicating malignant neoplasm, remove patients with diagnosis.

```{r}
unique(icd[grepl("^14\\d", icd$DIAGNOSIS_ICD9_CODE),]$DIAGNOSIS_LONG_TITLE)

tmp <- dat[!(dat$HADM_ID %in% icd[grepl("^14\\d", icd$DIAGNOSIS_ICD9_CODE),]$HADM_ID),]

cat(length(unique(dat$SUBJECT_ID)) - length(unique(tmp$SUBJECT_ID)), "patients removed.\n")
cat(length(unique(dat$HADM_ID)) - length(unique(tmp$HADM_ID)), "hospital admissions removed.\n")
cat(length(unique(dat$ICUSTAY_ID)) - length(unique(tmp$ICUSTAY_ID)), "ICU stays removed.\n")

## dat gets tmp, remove tmp
dat <- tmp
rm(tmp)
```

### Myasthinia gravis

Look for `35XX` ICDs indicating myasthinia gravis, remove patients with diagnosis.

```{r}
for (code in c("^3589", "^3590", "^3591", "^35801")) cat(gsub("\\^", '', code),": ", unique(icd[grepl(code, icd$DIAGNOSIS_ICD9_CODE),]$DIAGNOSIS_LONG_TITLE),'\n', sep = '')
rm(code)

## Subsetting
tmp <- dat[!(dat$HADM_ID %in% icd[grepl("^3589", icd$DIAGNOSIS_ICD9_CODE),]$HADM_ID),]
tmp <- tmp[!(tmp$HADM_ID %in% icd[grepl("^3590", icd$DIAGNOSIS_ICD9_CODE),]$HADM_ID),]
tmp <- tmp[!(tmp$HADM_ID %in% icd[grepl("^3591", icd$DIAGNOSIS_ICD9_CODE),]$HADM_ID),]
tmp <- tmp[!(tmp$HADM_ID %in% icd[grepl("^35801", icd$DIAGNOSIS_ICD9_CODE),]$HADM_ID),]

## Check
cat(length(unique(dat$SUBJECT_ID)) - length(unique(tmp$SUBJECT_ID)), "patients removed.\n")
cat(length(unique(dat$HADM_ID)) - length(unique(tmp$HADM_ID)), "hospital admissions removed.\n")
cat(length(unique(dat$ICUSTAY_ID)) - length(unique(tmp$ICUSTAY_ID)), "ICU stays removed.\n")

## dat gets tmp, remove tmp
dat <- tmp
rm(tmp)
```

### Burns

1. View list of burn diagnoses `94XX` ICDs.
2. Remove patients with diagnoses of burns to H/F/N or >30% BSA.

```{r}
## (1) View list of burn types
unique(icd[grepl("^94\\d", icd$DIAGNOSIS_ICD9_CODE),]$DIAGNOSIS_LONG_TITLE)

## Only H/F/N or >30% BSA
(codes <- unique(icd[grepl("^94\\d", icd$DIAGNOSIS_ICD9_CODE),]$DIAGNOSIS_LONG_TITLE)[c(1, 2, 3, 4, 5,
                                                                                        33, 34, 37, 38)])

## Find associated HADM_IDs
codes <- unique(icd[(icd$DIAGNOSIS_LONG_TITLE %in% codes),]$HADM_ID)

## (2) Removal
tmp <- dat[!(dat$HADM_ID %in% codes),]

## Numbers
cat(length(unique(dat$SUBJECT_ID)) - length(unique(tmp$SUBJECT_ID)), "patients with burns removed.\n")
cat(length(unique(dat$HADM_ID)) - length(unique(tmp$HADM_ID)), "hospital admissions associated with burn victims removed.\n")
cat(length(unique(dat$ICUSTAY_ID)) - length(unique(tmp$ICUSTAY_ID)), "ICU stays associated with burn victims removed.\n")

## dat gets tmp, remove tmp
dat <- tmp
rm(tmp, codes)
```

### Organ Transplant Patients

Remove organ donors as they could be cadaveric.

```{r}
tmp <- dat[!(dat$HADM_ID %in% icd[grepl("donor", icd$DIAGNOSIS_LONG_TITLE, ignore.case = TRUE),]$HADM_ID),]

## Numbers
cat(length(unique(dat$SUBJECT_ID)) - length(unique(tmp$SUBJECT_ID)), "patients removed as being organ donors.\n")
cat(length(unique(dat$HADM_ID)) - length(unique(tmp$HADM_ID)), "hospital admissions associated with organ donors.\n")
cat(length(unique(dat$ICUSTAY_ID)) - length(unique(tmp$ICUSTAY_ID)), "ICU stays associated with organ donors.\n")

## dat gets tmp, remove tmp
dat <- tmp
rm(tmp, icd)
```

## Check Ventilator Use

1. Load `ventdurations` (derived table)
2. Merge data, thus removing any patients which have no ventilation events

```{r}
## (1)
## ventdurations
vent <- read.csv("~/MIMIC-III/ventdurations.csv", header = T, stringsAsFactors = F)
## Convert column names to uppercase
colnames(vent) <- toupper(colnames(vent))

## (2)
## Remove patients with no ventilation data
tmp <- merge(dat, vent, by = "ICUSTAY_ID")

## Print numbers
cat(length(unique(dat$SUBJECT_ID)) - length(unique(tmp$SUBJECT_ID)), "patients removed for never being on a ventilator.\n")
cat(length(unique(dat$HADM_ID)) - length(unique(tmp$HADM_ID)), "hospital admissions removed for not being associated with ventilator use.\n")
cat(length(unique(dat$ICUSTAY_ID)) - length(unique(tmp$ICUSTAY_ID)), "ICU stays removed for not being associated with ventilator use.\n")

## dat gets tmp, remove tmp
dat <- tmp
rm(tmp, vent)

## Print
cat("There are", length(unique(dat$SUBJECT_ID)), "unique patients in this cohort.\n")
cat("There are", length(unique(dat$HADM_ID)), "unique hospital admissions in this cohort.\n")
cat("There are", length(unique(dat$ICUSTAY_ID)), "unique ICUSTAYs in this cohort.\n")


## Write csv
## write.csv(dat, "~/LindvallLab/trach_study_v2/trach_study_all20Sep19.csv", row.names = F)
```

### Time and Ventilator Selection

1. Convert data from character to dates for ordering
2. Order based on `SUBJECT_ID`, `HADM_ID`, `ICUSTAY_ID`, `INTIME`, `STARTTIME`
3. Create d7 cohort
4. Create d14 cohort
5. Add Cohort information and merge

```{r}
## (1) Vent time data
dat$STARTTIME <- strptime(dat$STARTTIME, "%Y-%m-%d %H:%M:%S", tz = "EST")
dat$ENDTIME <- strptime(dat$ENDTIME, "%Y-%m-%d %H:%M:%S", tz = "EST")
dat$ADMITTIME <- strptime(dat$ADMITTIME, "%Y-%m-%d %H:%M:%S", tz = "EST")
dat$INTIME <- strptime(dat$INTIME, "%Y-%m-%d %H:%M:%S", tz = "EST")
dat$OUTTIME <- strptime(dat$OUTTIME, "%Y-%m-%d %H:%M:%S", tz = "EST")
dat$DOD <- strptime(dat$DOD, "%Y-%m-%d %H:%M:%S", tz = "EST")
dat$DISCHTIME <- strptime(dat$DISCHTIME, "%Y-%m-%d %H:%M:%S", tz = "EST")

## Days Until Death
dat$DAYS_UNTIL_DEATH <- ifelse(is.na(dat$DOD), NA, as.numeric(dat$DOD - dat$ADMITTIME, units = "days"))

## (3)
tmpd7 <- ventCheck(dat, 7, "ICUSTAY_ID", 0.25)

## (4)
tmpd14 <- ventCheck(dat, 14, "ICUSTAY_ID", 0.25)

## (5)
tmpd7$COHORT <- rep("d7", each = nrow(tmpd7))
tmpd14$COHORT <- rep("d14", each = nrow(tmpd14))
```

### Check Cohort

```{r}
## Check
cat(length(unique(dat$SUBJECT_ID)) - length(unique(tmpd7$SUBJECT_ID)), "d7 patients removed.\n")
cat(length(unique(dat$HADM_ID)) - length(unique(tmpd7$HADM_ID)), "d7 hospital admissions removed.\n")
cat(length(unique(dat$ICUSTAY_ID)) - length(unique(tmpd7$ICUSTAY_ID)), "d7 ICU stays removed.\n")

## Check
cat(length(unique(dat$SUBJECT_ID)) - length(unique(tmpd14$SUBJECT_ID)), "d14 patients removed.\n")
cat(length(unique(dat$HADM_ID)) - length(unique(tmpd14$HADM_ID)), "d14 hospital admissions removed.\n")
cat(length(unique(dat$ICUSTAY_ID)) - length(unique(tmpd14$ICUSTAY_ID)), "d14 ICU stays removed.\n")

print_info(tmpd7)
print_info(tmpd14)
```

## Write data

```{r}
## Add cohort info
dat$COHORT <- rep("overall", each = nrow(dat))

write.csv(dat, file = "~/LindvallLab/trach_study_v2/entire_cohort27Sep19.csv", row.names = F)
write.csv(tmpd7, file = "~/LindvallLab/trach_study_v2/d7_mechvent_cohort27Sep19.csv", row.names = F)
write.csv(tmpd14, file = "~/LindvallLab/trach_study_v2/d14_mechvent_cohort27Sep19.csv", row.names = F)
```